\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-01-24 paper/main.tex
entered CEF and DR comments

We also designed two conditions where we varied the words' visual appearnaces
across lists. In the \textit{reduced (early)} condition, we followed the
``reduced'' procedure (presenting each word in black at the center of the
screen) for early lists, and followed the ``feature rich'' procedure
(presenting each word in a random color and location) for late lists. Finally,
in the \textit{reduced (late)} condition, we followed the feature rich
procedure for early lists and the reduced procedure for late lists.

@@

We found that \textit{adding} irrelevant visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature rich versus reduced (early):
$t(107) = -2.230, p = 0.028$; reduced versus reduced (early): $t(101) = -2.045,
p = 0.043$; also see Fig.~\accuracyByList A). However, \textit{subtracting}
irrelevant visual features on later lists that \textit{had} been present on
early lists (as in the reduced (late) condition) did not appear to impact
recall performance (accuracy for feature rich versus reduced (late): $t(106) =
-0.638, p = 0.525$; reduced versus reduced (late): $t(100) = -0.407, p =
0.685$). These comparisons suggest that recall accuracy has a directional
component: accuracy is affected differently by removing features later
that had been present earlier versus adding features later that had
\textit{not} been present earlier. In contrast, we found that participants
exhibited more temporal and feature-based clustering when we added irrelevant
visual features to \textit{any} lists (comparisons of clustering on feature
rich anversusd reduced lists are reported above; temporal clustering in reduced
versus reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-9.780$, $p$s $< 0.001$; feature-based clustering in reduced versus reduced
(early) and reduced versus reduced (late) conditions: $t$s $\leq -5.443$, $p$s
$< 0.001$). Temporal and feature-based clustering were not reliably different
in the feature rich, reduced (early), and reduced (late) conditions (temporal
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.434$, $p$s $\geq 0.154$; feature-based
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.359$, $p$s $> 0.177$).

@@

2.099, p = 0.038$; location: $t(126) = 4.392, p < 0.001$), but showed now

@@

We suggest two potential interpretations of these findings. First, it is
possible that some participants are more ``malleable'' or ``adaptable'' with
respect to how they organize incoming information. When presented with list of
items sorted along \textit{any} feature dimension, they will simply adopt that
feature as a dominant dimension for organizing those items and subsequent
(randomly ordered) items. This flexibility in memory organization might afford
such participants a memory advantage, explaining their strong recall
performance. An alternative interpretation is that each participant comes into
our study with a ``preferred'' way of organizing incoming information. If they
happen to be assigned to an order manipulation condition that matches their
preferences, then they will appear to be ``sensitive'' to the order
manipulation and also exhibit a high degree of carryover in feature clustering
from early to late lists. These participants might demonstrate strong recall
performance not because of their inherently superior memory abilities, but
rather because the specific condition they were assigned to happened to be
especially easy for them, given their pre-experimental tendencies. To help
distinguish between these interpretations, we designed an \textit{adaptive}
experimental condition (see \textit{Adaptive condition}). The primary
manipulation in the adaptive condition is that participants each experience
three key types of lists. On \textit{random} lists, words are ordered randomly
(as in the feature rich condition). On \textit{stabilize} lists, the
presentation order is adjusted to be maximally similar to the current estimate
of the participant's memory fingerprint (see \textit{Online “fingerprint”
analysis}). Third, on \textit{destabilize} lists, the presentation is adjusted
to be \textit{minimally} similar to the current estimate of the participant's
memory fingerprint (see \textit{Ordering ``stabilize'' and ``destabilize''
lists by an estimated fingerprint}). The orders in which participants
experienced each type of list were counterbalanced across participants to help
reduce the influence of potential list order effects. Because the presentation
orders on stabilize and destabilize lists are adjusted to best match each
participant's (potentially unique) memory fingerprint, the adaptive condition
removes uncertainty about whether participants' assigned conditions might just
``happen'' to match their preferred ways or organizing their memories.

>>>>

We also designed two conditions where we varied the words' visual appearances
across lists. In the \textit{reduced (early)} condition, we followed the
``reduced'' procedure (presenting each word in black at the center of the
screen) for early lists, and followed the ``feature rich'' procedure
(presenting each word in a random color and location) for late lists. Finally,
in the \textit{reduced (late)} condition, we followed the feature rich
procedure for early lists and the reduced procedure for late lists.

@@

We found that \textit{adding} irrelevant visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature rich versus reduced (early):
$t(107) = -2.230, p = 0.028$; reduced versus reduced (early): $t(101) = -2.045,
p = 0.043$; also see Fig.~\accuracyByList A). However, \textit{subtracting}
irrelevant visual features on later lists that \textit{had} been present on
early lists (as in the reduced (late) condition) did not appear to impact
recall performance (accuracy for feature rich versus reduced (late): $t(106) =
-0.638, p = 0.525$; reduced versus reduced (late): $t(100) = -0.407, p =
0.685$). These comparisons suggest that recall accuracy has a directional
component: accuracy is affected differently by removing features later
that had been present earlier versus adding features later that had
\textit{not} been present earlier. In contrast, we found that participants
exhibited more temporal and feature-based clustering when we added irrelevant
visual features to \textit{any} lists (comparisons of clustering on feature
rich versus reduced lists are reported above; temporal clustering in reduced
versus reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-9.780$, $p$s $< 0.001$; feature-based clustering in reduced versus reduced
(early) and reduced versus reduced (late) conditions: $t$s $\leq -5.443$, $p$s
$< 0.001$). Temporal and feature-based clustering were not reliably different
in the feature rich, reduced (early), and reduced (late) conditions (temporal
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.434$, $p$s $\geq 0.154$; feature-based
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.359$, $p$s $> 0.177$).

@@

2.099, p = 0.038$; location: $t(126) = 4.392, p < 0.001$), but showed no

@@

We suggest two potential interpretations of these findings. First, it is
possible that some participants are more ``malleable'' or ``adaptable'' with
respect to how they organize incoming information. When presented with list of
items sorted along \textit{any} feature dimension, they will simply adopt that
feature as a dominant dimension for organizing those items and subsequent
(randomly ordered) items. This flexibility in memory organization might afford
such participants a memory advantage, explaining their strong recall
performance. An alternative interpretation is that each participant comes into
our study with a ``preferred'' way of organizing incoming information. If they
happen to be assigned to an order manipulation condition that matches their
preferences, then they will appear to be ``sensitive'' to the order
manipulation and also exhibit a high degree of carryover in feature clustering
from early to late lists. These participants might demonstrate strong recall
performance not because of their inherently superior memory abilities, but
rather because the specific condition they were assigned to happened to be
especially easy for them, given their pre-experimental tendencies. To help
distinguish between these interpretations, we designed an \textit{adaptive}
experimental condition (see \textit{Adaptive condition}). The primary
manipulation in the adaptive condition is that participants each experience
three key types of lists. On \textit{random} lists, words are ordered randomly
(as in the feature rich condition). On \textit{stabilize} lists, the
presentation order is adjusted to be maximally similar to the current estimate
of the participant's memory fingerprint (see \textit{Online “fingerprint”
analysis}). Third, on \textit{destabilize} lists, the presentation is adjusted
to be \textit{minimally} similar to the current estimate of the participant's
memory fingerprint (see \textit{Ordering ``stabilize'' and ``destabilize''
lists by an estimated fingerprint}). The orders in which participants
experienced each type of list were counterbalanced across participants to help
reduce the influence of potential list order effects. Because the presentation
orders on stabilize and destabilize lists are adjusted to best match each
participant's (potentially unique) memory fingerprint, the adaptive condition
removes uncertainty about whether participants' assigned conditions might just
``happen'' to match their preferred ways of organizing their memories.
